Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2024-02-28 revision/review_response.qmd

---
format: 
  docx:
    reference-doc: response_styles.docx
bibliography: ../refs.bib
csl: ../limnology-and-oceanography.csl

@@

execute: 
  echo: false
  warning: false
  message: false
---

```{r echo = F, cache = F, eval = F}
#| eval: false
spelling::spell_check_files(here::here('revision/review-response.qmd'))
```

*We thank all three reviewers for providing thorough and constructive comments that have helped to improve the manuscript. Our responses to each of these comments are provided below.  Note that line numbers refer to those in the original manuscript.*

@@

The authors describe pitfalls in their model, particularly when it comes to estimating the gas transfer velocity. Additionally, the authors provide evidence that sometimes the right answer (i.e., modeled DO time series, lines 349-352) can be achieved through incorrect processes. Similar issues have been described and addressed in the stream metabolism literature which was not cited or discussed in this study, please see Appling et al. 2018 - https://doi.org/10.1002/2017JG004140). Consider adding to the discussion how EBASE might incorporate these advancements to address these issues.

*Response: text goes here*

Detiding (i.e., removing the influence of tidal cycles on time series data) the DO time series substantially improved model estimates and therefore was recommended as a step prior to applying the model. However, it was not clear how the authors detided their time series (they used Weighted Regression but it is unclear how this method removes tidal influence from the predicted time series – see lines 400-402). Please add additional information on the detiding approach and which variables were detided (i.e., was only DO detided? Wouldn’t all water-related input variables, such as water temperature and salinity, need to be detided?).

*Response: The detiding approach used herein is thoroughly described in another L&O paper cited in the main text [@Beck15]. Describing this approach in detail is beyond the scope of the paper, although we have added some text for clarification.  Also, the method has only been applied to dissolved oxygen data, so we have not used it for any of the remaining parameters. Clarification for both points has been added to line 399: "The results of this analysis return a predicted and detided dissolved oxygen time series, where the detided results are simply the predictions at average tidal height [see @Beck15 for details]. This method was only applied to dissolved oxygen since it has not been evaluated with any other parameters."*

@@

How would your results be impacted by using different observed data collection frequencies such as every 5-min, 15-min, 1 hour?

*Response: The resolution of "continuous" monitoring data has not been evaluated in this study, although our guess is that more frequent sampling may produce more robust estimates.  However, this may not matter given the time scales at which metabolism changes (i.e., > 1 day).  In our experience, estimates from the Odum open-water method are nearly identical using 15-minute or hourly samples.  Not sure the same would apply for EBASE given the fundamental differences between the approaches. We have added a point in the discussion on line 541 that speaks to this question: "These hypotheses may also consider frequency of continuous monitoring (e.g., 15 minute vs hourly) and how this may influence the results, as this is currently unknown for EBASE."*

@@

In my use of Bayesian models for estimating metabolism, with each model run you can get different metabolism estimates. This is due to the prior distribution space that the model can operate in. Is this something you have noticed with your model? If so, could you add a sentence somewhere explaining that a user should expect different modeled estimates with each model run?

*Response: Surprisingly, we have not encountered this issue with EBASE.  Obviously the results depend on the prior distributions, which was the purpose of our sensitivity analysis, although we have not noticed slightly different results for different runs with the same priors.  It is possible these differences are present, but small enough not to notice.  We would hope that a user of this approach would be familiar enough with Bayesian and MCMC methods to expect these minor differences.*

@@

*Response: Yes, C represents observed dissolved oxygen concentration (as noted on line 132).  We have retained C instead of using Cobs for simplicity in the model equations.  Further, the addition of the model input/output table as requested by another reviewer helps clarify the input data.*

Line 181: Specifying the model optimization period is an advancement of the BASE model but is not described in text. What does ‘optimization’ mean? What are the implications of using 1 day, 7 days, 30 days, or the entire time series length on the output? If you use 7 days, do you only get weekly estimates of metabolism?

*Response: This was clarified in the text.*

*Line 181: "First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$).  The optimization period determines how much data are used to estimate each parameter. Only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously given the dependence on other inputs that can vary at each time step (e.g., $PAR$ and wind). The optimization period can vary from a minimum of 1 day (defined as midnight to midnight the following day) to a maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day.  Sensitivity of the EBASE method to different optimization periods is described below."*

@@

*Line 555: "More conservative gaps can be chosen if, for example, semidiurnal tidal signals operating at time scales less than 12 hours are expected to influence the results.  However, detiding is strongly encouraged in these cases and the default gap size of 12 hours that describes the diurnal oxygen signal is likely appropriate after detiding."*

*Also note that data for an entire optimization period are removed if any gaps greater than the chosen gap size are present, providing an additional safety measure that the results are not based on inaccurately interpolated inputs.*

Line 188: With your improvements to the original BASE method, could EBASE be used to estimate metabolism in non-tidal, lotic ecosystems?

*Response: In theory, EBASE could be used in non-tidal, lotic systems, since the primary difference with BASE is the use of a wind- and temperature-based gas exchange parameterization appropriate for open-water systems [@Wanninkhof14].  However, additional comparisons are needed (e.g., Odum vs EBASE in a lake) and our thorough justification for the default prior distributions are based on the estuarine literature.  This is still a valuable poin and we have added text to line 519:*

*"Further, EBASE could be applied to other non-tidal, lotic systems, although additional comparisons with existing methods are needed and the default uninformed prior distributions should be chosen carefully given the justification herein based primarily on estuarine literature [e.g., @caffrey2004]."*

@@

Line 267: Consider rephrasing – “…has a structure similar to, yet noisier than, that of light efficiency…”.

*Response: Rephrased.*

Line 270-271: Wouldn't one expect the opposite? During periods of high productivity, ingassing should occur given the relatively high(er) rates of primary production. Then during periods of low productivity (i.e., outside of the growing season) that outgassing would dominate as respiration of allochthonous material continues to occur while primarily production would be light limited (assuming the example dataset used is in a temperate environment).

*Response: Yes, this is correct and can even be seen in Figure 2, e.g., late September to early October shows P > R with negative D (ingassing), then P < R with positive D (outgassing).  We have corrected this error in the text.*

@@

Line 276-278: As pointed to earlier in the manuscript, it is unclear what is meant by the optimization period. Why are the results in Figure 3 averaged by the optimization period? Why not show the daily estimates? I liked seeing the fit of the EBASE estimates on the synthetic time series however, I would like to see the fit of a linear regression of the synthetic (observed) estimates vs. the EBASE (predicted) estimates.

*Response: Please see the response to line 181. In particular, our revision notes that only one value each for $a$, $R$, and $b$ is returned for each optimization period. The vertical lines for each point in Figure 3 are credible intervals showing uncertainy around each individual estimate.  The statement in the caption that "the results are averaged at the time step of the model optimization period" is incorrect and removed.  This really only applies to the synthetic results in Figure 5 to show a more clear comparison with the recovered values from EBASE.*

@@

Line 313: This explanation of optimization needs to come much earlier in the text and be expanded upon.

*Response: Agreed, please see the response to line 181. This section was also slightly shortened in response to the changes on line 181.*

@@

Line 324-327: I'm not sure if I follow correctly - are you seeking daily metabolic estimates? If so, are the 7 & 30 day optimizations using a forward looking moving window approach that advances by one-day (similar to a rolling average)?

*Response: The text was edited for clarity: "Results from optimization periods of 1, 7, and 30 days were compared to the synthetic one-year time series ($b$ was fixed for the sensitivity analysis) based on daily averages for both the synthetic and EBASE results.  For the latter, parameters that are only returned as individual values for each optimization period ($a$, $R$, and $b$) were assigned as daily values for comparability of summary statistics where $n$ is the same across analyses."*

@@

Line 473-477: It is stated that “Bayesian…methods have not been rigorously explored for estimating metabolic rates in estuaries” however, I point back to the two cited studies in the introduction (line 98) that did use Bayesian techniques to estimate estuarine metabolism. Consider rewording this sentence and/or expand on why those earlier studies are not considered rigorous.

*Response: text goes here*

Line 581: Maybe I missed it but how many MCMC simulations were used? How would a user change the number of MCMC simulations in the EBASE package? Can you provide a recommendation for the number of simulations to use or at least how many were used in this analysis?

*Response: This detail was omitted and we have now included it on line 199 with the description of the MCMC approach (default 10,000 simulations).*

@@

Table of model parameters. There are several parameters, most are relatively simple (single letter), and they appear on figures. I think it would be valuable to have a table. This could include, name, acronym, units, and any other relevant information (source/equation/etc.). I can also see organizing this table by inputs and outputs (L146) to help future users identify their input data and expected outputs.

*Response: We have added a new table showing model inputs/outputs, parameter names, acronyms, and units.*

Is EBASE only applicable to estuaries? It seems that the main change from BASE is the inclusion of a wider temporal window to estimate model parameters and changes in the b parameterization. Can these modifications also improve upon metabolism estimates in other systems (lakes, streams, etc.)? While I see the goal of improving metabolism in estuaries, there may be wider applicability to rivers, lakes, ocean, or other coastal areas. If so, mentioning what kinds of systems (big lakes, deep lakes, big rivers, complex hydrodynamics, etc.?) would be most appropriate for the model and potential get improved estimates over other metabolism approaches. This may expand the impact of this paper.

*Response: Yes, it is possible with caveats.  Please see our response to reviewer on line 188.  Specifically, additional comparisons are needed and the default uninformed priors based on estuarine literature must be reconsidered.*

@@

L258: typo in “estimatd”

*Response: Fixed.*

L269: Should it be “availability” rather than “variability”? of course, a more variable pool of organic matter could also be correct.

*Response: Changed to "availability", this seems more correct.*

### 2024-02-28 manu-draft.qmd
more revisions

For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>). EBASE differs from BASE in model implementation in two ways. First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$, see below). The time period can vary from a minimum of 1 day (defined as midnight to midnight the following day) to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described below. Note that only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously with each timestep given the dependence on other inputs that can vary across the time series (e.g., $PAR$ and wind). Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. All gaps are linearly interpolated to allow the model estimation to proceed continuously and any results for an optimization period with gaps larger than 12 hours are removed, eliminating the need to pre- and post-process the data when using EBASE. The user can vary the criteria for removing results (e.g., 6 hours as a more conservative criteria). Again, this feature was not included in the original BASE method and a complete time series was required for use. A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. All results provided herein were generated using the EBASE R package.

@@

The parameters $a$, $R$, and $b$ are estimated by likelihood given the observed data and prior distributions for the parameters. The "Just Another Gibbs Sampler" (JAGS) software [as for BASE, @Plummer03] is used with the EBASE R package (described above) to estimate probability distributions of the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework applied to a finite-difference form of @eq-ebase. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed. All parameter inputs and outputs for EBASE are shown in @tbl-ebaseiotab.

@@

The synthetic oxygen time series was generated using @eq-ebase as a forward calculating model with specified daily values for $a$, $R$, and $b$. The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Daily values of $a$ and $R$ were estimatd by (1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck23a], (2) computing monthly averages of the resulting daily estimates of $P$ and $R$, (3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and (4) by fitting a spline function to the monthly averages of $R$ and $a$.

The resulting synthetic time series of light efficiency, dissolved oxygen budget terms ($P$, $D$, and $R$), and dissolved oxygen concentration are shown in @fig-synapa. The light efficiency has a complex structure, with maxima in the summer, fall, and winter, and a strong minimum in spring (Figure [-@fig-synapa]a). The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to, yet noisier than, that of light efficiency, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter variability and high temperature; and (3) the air--water gas flux mimic productivity on short time scales, with outgassing during high productivity and ingassing during low productivity. The synthetic dissolved oxygen time series shows high variability at multiple time scales (Figure [-@fig-synapa]c). There is a prominent diurnal cycle that is superimposed on subseasonal variability, presumably driven by radiation and winds.

@@

The optimization period used for a time series determines how much data are used to estimate the $a$ and $R$ parameters. Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being 1 day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$ and $R$ are returned by the model ($b$ is also returned, but was fixed for the sensitivity analysis). A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if biogeochemical variation occurs at time scales less than the optimization period. Results from optimization periods of 1, 7, and 30 days were compared for the synthetic one-year time series. For the models with 7 and 30 day optimization periods, the final period in the annual time series was excluded from the results due to incomplete number of observations compared to earlier periods (i.e., the total number of days in the time series was not evenly divided by 7 or 30, resulting in a remainder for the last period). Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.

@@

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The noise estimates were created using a weighted regression analysis [@Beck15; @Beck23a] on observed dissolved oxygen time series for Apalachicola Bay in 2021. This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series. Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were derived from the observed, predicted and detided time series. The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series (Figure [-@fig-synapanos]a). Both the residual and tidal noise were added to the original synthetic time series in Figure [-@fig-synapa]c and are shown in Figure [-@fig-synapanos]b. EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series. The default priors (Equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period. The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.

@@

As described in the materials and procedures section, EBASE is an estuarine adaptation of BASE [@Grace15], which was developed for stream applications and some assumptions of the underlying model of the latter were inappropriate for the former. The most critical change made to BASE was adopting the gas transfer velocity parameterization in @Wanninkhof14, which is primarily wind-based and more appropriate than the depth-dependent formulation in BASE. EBASE also removed the temperature dependency of respiration, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).

EBASE also provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23]. Most importantly, EBASE was developed to allow different model optimization periods, whereas BASE was developed to estimate results for a single day. Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters. Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., ecological and biogeochemical characteristics influencing the $a$ and $R$ parameters (such as community composition and nutrient availability) may persist for several days. Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period. However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location. Our example using the synthetic time series from Apalachicola Bay suggested that the most precise results were obtained using the 30 day optimization period, although results are likely to vary across systems. For most applications, very short (e.g., 1 day) or very long (e.g., \> 30 days) are not recommended, such that the former may produce unstable results and the latter may miss important events in the time series producing parameters that are overly generalized. However, an advantage of the Bayesian approach is the ability to evaluate different optimization periods and, therefore, the amount of data on which to estimate key parameters. Hypotheses can be generated on the amount of time over which critical metabolic parameters may vary across systems.

@@

![Comparison of EBASE results using default prior distributions to a synthetic time series created with known model parameters ($a$, $R$, and $b$). Results are compared for EBASE parameters estimated with a (a) 7 day and (b) 30 day optimization period. 95% credible intervals from the posterior distributions of the parameter estimates from EBASE are also shown. The results are averaged at the time step of the model optimization period.](figs/defplo.png){#fig-defplo}

>>>>

For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>). EBASE differs from BASE in model implementation in two ways. First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$). The optimization period determines how much data are used to estimate each parameter. Only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously given the dependence on other inputs that can vary at each time step (e.g., $PAR$ and wind). The optimization period can vary from a minimum of 1 day (defined as midnight to midnight the following day) to a maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described below. Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. All gaps are linearly interpolated to allow the model estimation to proceed continuously and any results for an optimization period with gaps larger than 12 hours are removed, eliminating the need to pre- and post-process the data when using EBASE. The user can vary the criteria for removing results (e.g., 6 hours as a more conservative criteria). Again, this feature was not included in the original BASE method and a complete time series was required for use. A detailed web page ([https://fawda123.github.io/EBASE/](#0){.uri}) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. All results provided herein were generated using the EBASE R package.

@@

The parameters $a$, $R$, and $b$ are estimated by likelihood given the observed data and prior distributions for the parameters. The "Just Another Gibbs Sampler" (JAGS) software [as for BASE, @Plummer03] is used with the EBASE R package (described above) to estimate probability distributions of the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations (default 10,000) in a Bayesian framework applied to a finite-difference form of @eq-ebase. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed. All parameter inputs and outputs for EBASE are shown in @tbl-ebaseiotab.

@@

The synthetic oxygen time series was generated using @eq-ebase as a forward calculating model with specified daily values for $a$, $R$, and $b$. The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Daily values of $a$ and $R$ were estimated by (1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck23a], (2) computing monthly averages of the resulting daily estimates of $P$ and $R$, (3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and (4) by fitting a spline function to the monthly averages of $R$ and $a$.

The resulting synthetic time series of light efficiency, dissolved oxygen budget terms ($P$, $D$, and $R$), and dissolved oxygen concentration are shown in @fig-synapa. The light efficiency has a complex structure, with maxima in the summer, fall, and winter, and a strong minimum in spring (Figure [-@fig-synapa]a). The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to, yet noisier than, that of light efficiency, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter availability and high temperature; and (3) the air--water gas flux mimic productivity on short time scales, with ingassing during high productivity and outgassing during low productivity. The synthetic dissolved oxygen time series shows high variability at multiple time scales (Figure [-@fig-synapa]c). There is a prominent diurnal cycle that is superimposed on subseasonal variability, presumably driven by radiation and winds.

@@

The optimization period used for a time series determines how much data are used to estimate the metabolic parameters. A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if biogeochemical variation occurs at time scales less than the optimization period. Results from optimization periods of 1, 7, and 30 days were compared to the synthetic one-year time series (\$b\$ was fixed for the sensitivity analysis) based on daily averages for both the synthetic and EBASE results. For the latter, parameters that are only returned as individual values for each optimization period (\$a\$, \$R\$, and \$b\$) were assigned as daily values for comparability of summary statistics where \$n\$ is the same across analyses. For the models with 7 and 30 day optimization periods, the final period in the annual time series was excluded from the results due to incomplete number of observations compared to earlier periods (i.e., the total number of days in the time series was not evenly divided by 7 or 30, resulting in a remainder for the last period). Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.

@@

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The noise estimates were created using a weighted regression analysis [@Beck15; @Beck23a] on observed dissolved oxygen time series for Apalachicola Bay in 2021. This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series, where the detided results are simply the predictions at average tidal height [see @Beck15 for details]. This method was only applied to dissolved oxygen since it has not been evaluated with any other parameters. Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were derived from the observed, predicted and detided time series. The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series (Figure [-@fig-synapanos]a). Both the residual and tidal noise were added to the original synthetic time series in Figure [-@fig-synapa]c and are shown in Figure [-@fig-synapanos]b. EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series. The default priors (Equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period. The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.

@@

As described in the materials and procedures section, EBASE is an estuarine adaptation of BASE [@Grace15], which was developed for stream applications and some assumptions of the underlying model of the latter were inappropriate for the former. The most critical change made to BASE was adopting the gas transfer velocity parameterization in @Wanninkhof14, which is primarily wind-based and more appropriate than the depth-dependent formulation in BASE. EBASE also removed the temperature dependency of respiration, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS). Further, EBASE could be applied to other non-tidal, lotic systems, although additional comparisons with existing methods are needed and the default uninformed prior distributions should be chosen carefully given the justification herein based primarily on estuarine literature [e.g., @caffrey2004].

EBASE also provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23]. Most importantly, EBASE was developed to allow different model optimization periods, whereas BASE was developed to estimate results for a single day. Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters. Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., ecological and biogeochemical characteristics influencing the $a$ and $R$ parameters (such as community composition and nutrient availability) may persist for several days. Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period. However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location. Our example using the synthetic time series from Apalachicola Bay suggested that the most precise results were obtained using the 30 day optimization period, although results are likely to vary across systems. For most applications, very short (e.g., 1 day) or very long (e.g., \> 30 days) are not recommended, such that the former may produce unstable results and the latter may miss important events in the time series producing parameters that are overly generalized. However, an advantage of the Bayesian approach is the ability to evaluate different optimization periods and, therefore, the amount of data on which to estimate key parameters. Hypotheses can be generated on the amount of time over which critical metabolic parameters may vary across systems. These hypotheses may also consider frequency of continuous monitoring (e.g., 15 minute vs hourly) and how this may influence the results, as this is currently unknown for EBASE.

@@

![Comparison of EBASE results using default prior distributions to a synthetic time series created with known model parameters ($a$, $R$, and $b$). Results are compared for EBASE parameters estimated with a (a) 7 day and (b) 30 day optimization period. 95% credible intervals from the posterior distributions of the parameter estimates from EBASE are also shown.](figs/defplo.png){#fig-defplo}
